Degrees Draft Syllabus

### 2023-03-23 Draft Syllabus.Rmd
finalizing lesson 3

During the course we are going to work primarily (exclusively?) in R and much of the work done using R will be done with R Markdown. Though there are many other languages used for data analysis (including Python, SAS, Stata, and SQL), we will not explore those. If you'd like to know more about one of them feel free to reach out to me and I am happy to share what I know or help point you to a starting place if I am inexperienced in the language myself. And it is okay to be inexperienced! Don't worry if you are not familiar with the language or with coding at all, this class and the Codecademy course that will supplement it are meant for beginners. If you do have experience with R and feel that the course is moving too slowly or that you are not being sufficiently challenged, we encourage you to 1) Add complexity to the examples and challenges presented in class or on Codecademy, 2) Visit some of the pages provided in the enrichment section of this syllabus, 3) read ahead in the syllabus and work on something that is at your level, or 4) focus extra time on your project.

@@

The course will be split into two units. Unit 1 is Learning. During the Learning Unit we will be focused on working through lessons. Each session will be split into two parts: 1) a fifteen minute mini-lesson delivered by me or another member of the #DegreesNYC team, 2) a Codecademy lesson. The mini-lessons are designed to align as closely as possible with the Codecademy lesson for the week to provide as much reinforcement and clarification on the topics that we will cover as possible.

@@

This syllabus will be the repository for everything course-related. Information about lessons, the project, and enrichment is all contained here. The table of contents in the top left has links to the content for units 1 and 2, the introduction, and enrichment activities. Within the Unit 1 header there are tabs for each of the XXXX lessons that we will work through during our time together. These lessons each also have tabs for the topics we will cover, the lesson itself, and any helpful links that you might want to explore. This document will be hosted publically on RPubs, but it is a living document. If anything changes or if there are problems with the URL leading to the syllabus, I will let you know as soon as possible.

@@

![R is a very common language in data science](./Lesson 1/r_prevalence.png)
R is one of the most common languages for data science and data analysis. This is because, with experience, one can take a project from beginning to end only using R. You can download data; manipulate it by creating new variables, dropping missing observations, etc.; perform statistical analysis; and visualize and output results. R, like Stata and SAS, has built-in statistical analysis functionality, but unlike those two languages it is 100% free and open source. There is a robust range of functions in base R, but the R community is very strong and has built a host of tools that make doing data analysis and visualization much easier than it is in base R (we will be working with one such suite of tools, the tidyverse/dplyr, in this class). If you are hoping to get a job doing analytic work, if you want advanced education in a social science field, if you want an introduction to coding/computer science, or if you are just looking to better understand quantitative analysis, R is a good place to begin.

@@

As noted in the introduction, this class will use Codecademy to aid in learning basic coding skills in R and the basic R workflow. You will need a Codecademy account to retain your progress between sessions. You can register for one [here](https://www.codecademy.com/)
<!-- Should I add in Git here? -->
Another tool we will use throughout the class and one you will also be using in your Codecademy work is a suite of R packages called the Tidyverse. The Tidyverse is designed to make working with data easier and is commonly used in many organizations (we use it at Mathematica, for example). There are eight major packages in the Tidyverse as well as many other specialised packages, but the following are important for this class:

@@

Finally, we will use R Markdown when creating your projects. R Markdown allows you to write code and text into the same document and have it rendered nicely as a word document, PDF, or html file. I used R Markdown to write this syllabus. Using R Markdown makes integrating visualizations, code, and analyses into your reports simple.

@@

While we are in class you can ask me questions and, of course, feel free to email me and other members of the #DegreesNYC team as you need support, but there are other things you can do to get answers before asking me or a classmate for help.

@@

Variables allow us to work with values dynamically and without the struggle of trying to remember all of the values that we are working with individually. Keeping the math theme, suppose you wanted to implement the quadratic formula which is written:<br/>
![](./Lesson 2/quadratic_formula.png)<br/>

@@

```{r l2_b_ex4}
a = 2
b = 4
2 -> c
(-b + sqrt(b^2 - 4*a*c))/(2*a)
```
The advantage of the second option is that if the numbers you would like to enter into formula change, all you have to do is change them once, when assigning them, and you can use the formula again. In the first case, you would need to go through and make sure you've got all of the numbers entered into their correct locations carefully. The true power of variables will not be seen until we begin to work with for loops, but, for now, think of them as useful ways to store information.

@@

In this class we will focus on the logical, numeric, and character data types. These are the three most common types. See the helpful links for more information about the types that we will not cover.

@@

The first line of code throws an error, telling us that "1" and "2" are not numeric and therefore cannot be added. The second line of code returns what we expect, 3.

@@

You are also likely familiar with most of the logical operators. These are things in math which do not return a number, but rather return a value of true or false. Greater than (>), less than (<), greater than or equal to (>=), less than or equal to (<=), equal to (==), and not equal to (!=) are all available in R. Note that you must use 2 equal signs in R when you what to equate two things. One equal sign, as we discussed earlier, is for assigning values to variables.

@@

* [You can read more about the data types that we do not cover here.](https://www.programiz.com/r/data-types)
* [More information about operators](https://www.statmethods.net/management/operators.html)

@@

One of R's strengths, that it is open source, is also what makes it difficult to learn. There are sometimes many ways to do the same thing with solutions coming from base R, as well as many different packages. For example, suppose we load the built-in R data set mtcars and want to view only cars with mpg > 20:

@@

@@

dplyr's filter function allows you to select rows meeting certain criteria. Suppose we are using the mtcars data set and only want to see cars with exactly 6 cylinders. Then with dplyr:

@@

Or suppose we only wanted cars with 4 or more forward gears. Then

@@

Combine filter with logical operators and other functions for more complex selection of observations. Like in this case where the data is subset to only cars made by Mazda:

@@

Slice allows selecting rows by index. For example, to select rows 1, 3, and 5 of mtcars:

@@

Arrange sorts data. You provide a column and, by default, it sorts the rows by that column in ascending order. Compare the following:

```{r l3_2_ex9}
mtcars
mtcars %>%
  arrange(mpg)
```

In the first case, the mtcars data set is in some arbitrary order. After using arrange, the rows are ordered from lowest to highest mpg. Wrapping the target columns in `desc()` will arrange rows in descending order:

@@

Providing multiple columns to arrange will break ties on preceding columns:

@@

Select is among the most common dplyr verbs. It has a simple, but important function: selecting which variables should be kept in a data frame. For example, if we only want to keep the mpg variable from the mtcars data frame, we can do the following:

@@

If instead, we'd like to keep multiple variables--say cylinders, gears, and carburetors:

@@

Sometimes column names are inconvenient to work with, not descriptive, or generally do not meet your needs. Rename can remedy this problem; use it when you want to rename a column in a data frame:

@@

R has rules about how variables can be named (see `help(make.names)` for more information). In general, variable names cannot have non-alphanumeric characters, cannot begin with numbers, and cannot contain spaces. Still, if you would really like to (for example if you are creating a public-facing table), you can create non-syntactic names by wrapping them in backticks:

@@

Mutate allows you to modify and create columns in a data frame. You use mutate by writing the column name you'd like to modify/create on the left-hand side and an expression on the right-hand side of an expression. For example we can calculate the mpg to cyl ratio in mtcars:

@@

As with rename, if we'd like to manipulate multiple variables in one call to mutate we can do that by separating expressions with a comma:

@@

###### **relocate**
###### **summarise**

>>>>

During the course you are going to work primarily (exclusively?) in R and much of the work done using R will be done with R Markdown. Though there are many other languages used for data analysis (including Python, SAS, Stata, and SQL), you will not explore those. If you'd like to know more about one of them feel free to reach out to me and I am happy to share what I know or help point you to a starting place if I am inexperienced in the language myself. And it is okay to be inexperienced! Don't worry if you are not familiar with the language or with coding at all, this class and the Codecademy course that will supplement it are meant for beginners. If you do have experience with R and feel that the course is moving too slowly or that you are not being sufficiently challenged, you encourage you to 1) Add complexity to the examples and challenges presented in class or on Codecademy, 2) Visit some of the pages provided in the enrichment section of this syllabus, 3) read ahead in the syllabus and work on something that is at your level, or 4) focus extra time on your project.

@@

The course will be split into two units. Unit 1 is Learning. During the Learning Unit you will be focused on working through lessons. Each session will be split into two parts: 1) a fifteen minute mini-lesson delivered by me or another member of the #DegreesNYC team, 2) a Codecademy lesson. The mini-lessons are designed to align as closely as possible with the Codecademy lesson for the week to provide as much reinforcement and clarification on the topics that you will cover as possible.

@@

This syllabus will be the repository for everything course-related. Information about lessons, the project, and enrichment is all contained here. The table of contents in the top left has links to the content for units 1 and 2, the introduction, and enrichment activities. Within the Unit 1 header there are tabs for each of the XXXX lessons that you will work through during our time together. These lessons each also have tabs for the topics you will cover, the lesson itself, and any helpful links that you might want to explore. This document will be hosted publically on RPubs, but it is a living document. If anything changes or if there are problems with the URL leading to the syllabus, I will let you know as soon as possible.

@@

![R is a very common language in data science](./Lesson 1/r_prevalence.png)
R is one of the most common languages for data science and data analysis. This is because, with experience, one can take a project from beginning to end only using R. You can download data; manipulate it by creating new variables, dropping missing observations, etc.; perform statistical analysis; and visualize and output results. R, like Stata and SAS, has built-in statistical analysis functionality, but unlike those two languages it is 100% free and open source. There is a robust range of functions in base R, but the R community is very strong and has built a host of tools that make doing data analysis and visualization much easier than it is in base R (you will be working with one such suite of tools, the tidyverse/dplyr, in this class). If you are hoping to get a job doing analytic work, if you want advanced education in a social science field, if you want an introduction to coding/computer science, or if you are just looking to better understand quantitative analysis, R is a good place to begin.

@@

As noted in the introduction, this class will use Codecademy to aid in learning basic coding skills in R and the basic R workflow. You will need a Codecademy account to retain your progress between sessions. You can register for one [here](https://www.codecademy.com/)
<!-- Should I add in Git here? -->
Another tool you will use throughout the class and one you will also be using in your Codecademy work is a suite of R packages called the Tidyverse. The Tidyverse is designed to make working with data easier and is commonly used in many organizations (you use it at Mathematica, for example). There are eight major packages in the Tidyverse as well as many other specialised packages, but the following are important for this class:

@@

Finally, you will use R Markdown when creating your projects. R Markdown allows you to write code and text into the same document and have it rendered nicely as a word document, PDF, or html file. I used R Markdown to write this syllabus. Using R Markdown makes integrating visualizations, code, and analyses into your reports simple.

@@

While you are in class you can ask me questions and, of course, feel free to email me and other members of the #DegreesNYC team as you need support, but there are other things you can do to get answers before asking me or a classmate for help.

@@

Variables allow us to work with values dynamically and without the struggle of trying to remember all of the values that you are working with individually. Keeping the math theme, suppose you wanted to implement the quadratic formula which is written:<br/>
![](./Lesson 2/quadratic_formula.png)<br/>

@@

```{r l2_b_ex4}
a = 2
b = 4
2 -> c
(-b + sqrt(b^2 - 4*a*c))/(2*a)
```
The advantage of the second option is that if the numbers you would like to enter into formula change, all you have to do is change them once, when assigning them, and you can use the formula again. In the first case, you would need to go through and make sure you've got all of the numbers entered into their correct locations carefully. The true power of variables will not be seen until you begin to work with for loops, but, for now, think of them as useful ways to store information.

@@

In this class you will focus on the logical, numeric, and character data types. These are the three most common types. See the helpful links for more information about the types that you will not cover.

@@

The first line of code throws an error, telling us that "1" and "2" are not numeric and therefore cannot be added. The second line of code returns what you expect, 3.

@@

You are also likely familiar with most of the logical operators. These are things in math which do not return a number, but rather return a value of true or false. Greater than (>), less than (<), greater than or equal to (>=), less than or equal to (<=), equal to (==), and not equal to (!=) are all available in R. Note that you must use 2 equal signs in R when you what to equate two things. One equal sign, as you discussed earlier, is for assigning values to variables.

@@

* [You can read more about the data types that you do not cover here.](https://www.programiz.com/r/data-types)
* [More information about operators](https://www.statmethods.net/management/operators.html)

@@

One of R's strengths, that it is open source, is also what makes it difficult to learn. There are sometimes many ways to do the same thing with solutions coming from base R, as well as many different packages. For example, suppose you load the built-in R data set mtcars and want to view only cars with mpg > 20:

@@

Two additional notes: 1) For folks who are coming from other oriented programming languages it might be useful to think of the pipe as similar to method chaining. The two are not identical, but the comparison may be useful for understanding why the pipe is worthwhile. 2) More recent versions of R have a built-in pipe which does not have to be loaded from magrittr `|>`. You can turn that on by going to Tools > Global Options > Code and selecting "Use native pipe operator".

@@

dplyr's `filter()` function allows you to select rows meeting certain criteria. Suppose you are using the mtcars data set and only want to see cars with exactly 6 cylinders. Then with dplyr:

@@

Or suppose you only wanted cars with 4 or more forward gears. Then

@@

Combine `filter()` with logical operators and other functions for more complex selection of observations. Like in this case where the data is subset to only cars made by Mazda:

@@

`slice()` allows selecting rows by index. For example, to select rows 1, 3, and 5 of mtcars:

@@

`arrange()` sorts data. You provide a column and, by default, it sorts the rows by that column in ascending order. Compare the following:

```{r l3_2_ex9}
# Not arranged
mtcars

# Data arranged by mpg
mtcars %>%
  arrange(mpg)
```

In the first case, the mtcars data set is in some arbitrary order. After using `arrange()`, the rows are ordered from lowest to highest mpg. Wrapping the target columns in `desc()` will arrange rows in descending order:

@@

Providing multiple columns to arrange will break ties on using succeeding columns:

@@

`select()` is among the most common dplyr verbs. It has a simple, but important function: selecting which variables should be kept in a data frame. For example, if you only want to keep the mpg variable from the mtcars data frame, you can do the following:

@@

If instead, you'd like to keep multiple variables--say cylinders, gears, and carburetors:

@@

Sometimes column names are inconvenient to work with, not descriptive, or generally do not meet your needs. `rename()` can remedy this problem; use it when you want to rename a column in a data frame:

@@

R has rules about how variables can be named (see `help(make.names)` for more information). If you would really like to (for example if you are creating a public-facing table), you can create non-syntactic names by wrapping them in backticks:

@@

`mutate()` allows you to modify and create columns in a data frame. You use mutate by writing the column name you'd like to modify/create on the left-hand side and a value on the right-hand side of an expression. For example you can calculate the mpg to cyl ratio in mtcars:

@@

As with `rename()`, if you'd like to manipulate multiple variables in one call to `mutate()` you can do that by separating expressions with a comma:

@@

Sometimes you want to apply the same transformation to multiple columns at once. There are several ways to do this using dplyr, but a common one is the function `across()`. `across()` is especially useful when summarizing, but can also be used in conjunction with mutate. For example, suppose you wanted to calculate the mean of cyl, mpg, and carb:

```{r l3_2_ex21}
mtcars %>%
  mutate(across(c(cyl, mpg, carb), mean, na.rm = T)) %>%
  select(mpg, cyl, carb)
```

This example is not particularly useful, as you would probably never want to imply that all car makes have the same mpg, cyl, and carb, but it illustrates the power of across.

###### **relocate**

`relocate()` allows you to move the location of columns in a data frame. You provide a series of columns that you would like to move and a column you would like to move them in front of (with `.before`) or behind (with `.after`). For example, to move the disp and hp columns before the mpg column you could do the following:

```{r l3_2_ex22}
mtcars %>%
  relocate(disp, hp, .before = mpg)
```

or

```{r l3_2_ex23}
mtcars %>%
  relocate(mpg, cyl, .after = hp)
```

`relocate()` is especially useful when outputting tables (where the order of variables is important).

###### **summarize**

`summarize()` (or `summarise()`) collapses data given a set of grouping variables created by `group_by()` or other means. The most common use of `summarize()` is for calculating summary statistics:

```{r l3_2_ex24}
# Group mtcars by mpg and calculate the mean cylinders and max carburetors within those groups.
mtcars %>%
  group_by(mpg) %>%
  summarize(cyl = mean(cyl),
            carb = max(carb))
```

But since `summarize()` applies functions within groups, it can be used for other things, such as grouping text:

```{r l3_2_ex25}
# Create mock data
ids <- c(1, 1, 2, 2)
text <- c("Hi", "I'm", "Reggie", "G.")
text_data <- as.data.frame(list(id = ids, text = text))
head(text_data)

# Combine text by ID
text_data %>%
  group_by(id) %>%
  summarize(text = paste(text, collapse = " "))
```

This can come in handy in a variety of contexts such as working with survey data or parsing a pdf.

#### **Tidy data**

In general, data sets that you come across while working or doing research will be messy. They will contain unexpected values, missingness, variables that are unclear, and other problems. To help make the irregularities of data more apparent and to shape data into a form that is easily analyzed, statisticians and programmers have created a set of rules for data collectively giving rise to the idea of tidy data. There are three important things to remember about tidy data (see `vignette("tidy-data")`, from which these rules are drawn):

1. Every column is a variable.
2. Every row is an observation.
3. Every cell is a single value.

As you begin to work with more sophisticated data sets, keep these rules in mind. Use the tools that you learn in this course and beyond to mold the data into a tidy data frame. Reshaping and massaging data may seem like time wasted working on something other than your analysis, but the time you spend organizing your data will be paid for by a more straightforward analytic process.
